\documentclass[12pt,a4paper]{article}

% -----------------------------------------------------------------------
% Packages
% -----------------------------------------------------------------------
\usepackage{amsmath,amsthm,amssymb}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{microtype}
\usepackage{enumitem}
\usepackage{url}
\usepackage{booktabs}

\geometry{margin=1in}

% -----------------------------------------------------------------------
% Theorem environments
% -----------------------------------------------------------------------
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% -----------------------------------------------------------------------
% Macros
% -----------------------------------------------------------------------
\newcommand{\floor}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\ceil}[1]{\left\lceil #1 \right\rceil}

% -----------------------------------------------------------------------
% Title
% -----------------------------------------------------------------------
\title{A Proof that $a(n)/n^2 \to \tfrac{1}{2}$\\
       for OEIS Sequence A112509}

\author{Ben Clare}

\date{April 2026}

% =======================================================================
\begin{document}
% =======================================================================

\maketitle

% -----------------------------------------------------------------------
\begin{abstract}
% -----------------------------------------------------------------------
For $n \ge 1$, let $a(n)$ denote the maximum number of distinct
non-negative integers whose binary representations appear as contiguous
substrings of some $n$-bit binary string (OEIS sequence A112509).
This note proves that
\[
  \lim_{n \to \infty} \frac{a(n)}{n^2} = \frac{1}{2}.
\]
The upper bound $a(n) \le n(n+1)/2$ is elementary.  For the matching
lower bound an explicit family of $n$-bit strings is exhibited achieving
at least $\frac{1}{2}n^2 - O(n^{5/3})$ distinct integer values.  The
argument is self-contained and elementary.
\end{abstract}

% =======================================================================
\section{Introduction}
% =======================================================================

Given a binary string $s$ of length $n$, every contiguous substring can
be read as a non-negative integer in binary (with leading zeros ignored
in the standard way: $\texttt{01} = \texttt{1} = 1$, $\texttt{00} =
\texttt{0} = 0$).

\begin{definition}
  For a binary string $s \in \{0,1\}^n$, let $f(s)$ denote the number
  of distinct non-negative integers whose binary representations appear
  among the contiguous substrings of $s$.  Define
  \[
    a(n) = \max_{s \in \{0,1\}^n} f(s).
  \]
\end{definition}

The sequence $a(n)$ is listed as OEIS A112509~\cite{oeis}.  It begins
\[
  1, 3, 5, 7, 10, 13, 17, 22, 27, 33, 40, 47, 55, 64, 73, 83, 94, 106,
  118, 131, \ldots
\]
Prior computational work~\cite{github} established exact values for
$n \le 150$ and certified lower bounds up to $n = 2 \times 10^9$ via
suffix-array-based search.  Selected values:

\begin{center}
\begin{tabular}{rrl}
  \toprule
  $n$ & $a(n)$ or lower bound & $a(n)/n^2 \ge$ \\
  \midrule
  50            & $899$                               & $0.3596$ \\
  100           & $3{,}875$                           & $0.3875$ \\
  150           & $9{,}070$                           & $0.4031$ \\
  $10^6$        & $498{,}396{,}662{,}200$             & $0.4984$ \\
  $10^9$        & $499{,}943{,}911{,}277{,}037{,}650$ & $0.49994$ \\
  $2\times10^9$ & $1{,}999{,}841{,}302{,}297{,}432{,}600$ & $0.49996$ \\
  \bottomrule
\end{tabular}
\end{center}

\noindent
The purpose of this note is to prove the limit rigorously.

\begin{theorem}\label{thm:main}
  $\displaystyle\lim_{n \to \infty} \dfrac{a(n)}{n^2} = \dfrac{1}{2}$.
\end{theorem}

The proof combines an elementary upper bound (Section~\ref{sec:upper})
with an explicit lower-bound construction (Sections~\ref{sec:construction}
--\ref{sec:asymptotic}).

% =======================================================================
\section{Upper bound}\label{sec:upper}
% =======================================================================

\begin{lemma}[Upper bound]\label{lem:upper}
  For all $n \ge 1$, $a(n) \le \dfrac{n(n+1)}{2}$.
\end{lemma}

\begin{proof}
  Two substrings of $s$ represent the same integer if and only if they
  are identical after stripping leading zeros.  Consequently every
  distinct positive integer value corresponds to a unique substring
  beginning with \texttt{1}.

  \emph{Case 1.}  Suppose $s$ contains at least one \texttt{0} bit.
  Then the value $0$ is also represented, but only by substrings beginning
  with \texttt{0}.  At least one of the $\frac{n(n+1)}{2}$ substrings
  begins with~\texttt{0}, so the count of leading-\texttt{1} substrings
  is at most $\frac{n(n+1)}{2} - 1$.  Adding the single value $0$ gives
  $f(s) \le \frac{n(n+1)}{2}$.

  \emph{Case 2.}  If $s$ is the all-ones string then $f(s)$ equals the
  number of distinct positive values, which is at most $\frac{n(n+1)}{2}$.

  Hence $a(n) \le \frac{n(n+1)}{2}$ and $\limsup_{n\to\infty}
  a(n)/n^2 \le \frac{1}{2}$.
\end{proof}

\begin{remark}
  The bound $a(n) \le n(n+1)/2$ is tight at $n = 2$: the string
  \texttt{11} has substrings \texttt{1}, \texttt{1}, \texttt{11} with
  values $1, 1, 3$, giving $f(\texttt{11}) = 2 = 2 \cdot 3/2 - 1$.
  Wait --- $\texttt{10}$ gives substrings with values $\{0, 1, 2\}$, so
  $a(2) = 3 = 2 \cdot 3/2$.  The bound is achieved.
\end{remark}

% =======================================================================
\section{The construction}\label{sec:construction}
% =======================================================================

Fix $n \ge 8$.  Set
\[
  m = \floor{n^{1/3}}, \qquad
  C = \frac{m(m-1)}{2}, \qquad
  L = \floor{\frac{n - C}{m}}, \qquad
  r = n - C - mL.
\]
Since $L = \floor{(n-C)/m}$ we have $0 \le r < m$ and
$n = mL + C + r$.  Note that $m \ge 2$ for $n \ge 8$, and
$L \ge (n-C)/m - 1 \ge n/(2m) \to \infty$, so $L \ge 1$ for all
sufficiently large $n$ (in fact for all $n \ge 8$).

Define the $n$-bit binary string
\begin{equation}\label{eq:sn}
  s_n \;=\;
  \underbrace{1^L}_{}\,
  0^1\,
  \underbrace{1^L}_{}\,
  0^2\,
  \underbrace{1^L}_{}\,
  0^3\,
  \cdots\,
  \underbrace{1^L}_{}\,
  0^{m-1}\,
  \underbrace{1^{L+r}}_{}
\end{equation}
consisting of $m$ one-blocks.  The first $m-1$ one-blocks each have
length $L$; the final one-block has length $L + r$.  The zero-separator
lengths are $1, 2, 3, \ldots, m-1$ in order.

\begin{lemma}[Length check]
  The string $s_n$ has length $n$.
\end{lemma}

\begin{proof}
  The total length is $(m-1)L + (L+r) + (1 + 2 + \cdots + (m-1))
  = mL + r + C = n$.
\end{proof}

% =======================================================================
\section{A large injective family of substrings}\label{sec:family}
% =======================================================================

For integers $i, j, \alpha, \beta$ satisfying
\[
  1 \le i < j \le m, \qquad 1 \le \alpha \le L, \qquad 1 \le \beta \le L,
\]
define the substring $u_{i,j,\alpha,\beta}$ of $s_n$ obtained by starting
inside the $i$-th one-block at the position where exactly $\alpha$ ones
remain in that block, and ending inside the $j$-th one-block immediately
after its first $\beta$ ones.

Because each intermediate one-block (blocks $i+1$ through $j-1$) has
length exactly $L$ and $\beta \le L \le L+r$ (so the endpoint lies inside
the $j$-th block regardless of whether $j = m$), the substring is
well-defined.  Explicitly, its binary form is
\begin{equation}\label{eq:u}
  u_{i,j,\alpha,\beta}
  \;=\;
  \underbrace{1^{\alpha}}_{\text{tail of block }i}\,
  \underbrace{0^{i}}_{\text{sep.}}\,
  \underbrace{1^{L}\,0^{i+1}}_{\text{block }i+1}\,
  \underbrace{1^{L}\,0^{i+2}}_{\text{block }i+2}\,
  \cdots\,
  \underbrace{1^{L}\,0^{j-1}}_{\text{block }j-1}\,
  \underbrace{1^{\beta}}_{\text{head of block }j}.
\end{equation}
Every such word begins and ends with \texttt{1}, so it is already the
standard binary expansion of a positive integer.

\begin{lemma}[Injectivity]\label{lem:inject}
  The map $(i, j, \alpha, \beta) \mapsto u_{i,j,\alpha,\beta}$ is
  injective: distinct quadruples yield distinct binary strings.
\end{lemma}

\begin{proof}
  Let $(i, j, \alpha, \beta)$ and $(i', j', \alpha', \beta')$ be two
  quadruples with $u_{i,j,\alpha,\beta} = u_{i',j',\alpha',\beta'}$ as
  binary strings.  Since both strings begin and end with \texttt{1}, their
  run-length encodings are uniquely determined.  From~\eqref{eq:u}, the
  run-length encoding of $u_{i,j,\alpha,\beta}$, listing alternating
  one-block and zero-block lengths, is
  \[
    (\alpha,\; i,\; L,\; i+1,\; L,\; i+2,\; \ldots,\; L,\; j-1,\; \beta).
  \]
  Equality of run-length encodings gives:
  \begin{itemize}
    \item First entry: $\alpha = \alpha'$.
    \item Last entry: $\beta = \beta'$.
    \item First zero-block length (second entry): $i = i'$.
    \item Last zero-block length: $j - 1 = j' - 1$, so $j = j'$.
  \end{itemize}
  Hence $(i, j, \alpha, \beta) = (i', j', \alpha', \beta')$, proving
  injectivity.
\end{proof}

% =======================================================================
\section{Lower bound and asymptotics}\label{sec:asymptotic}
% =======================================================================

\begin{lemma}[Counting]\label{lem:count}
  $a(n) \ge \dbinom{m}{2} L^2$.
\end{lemma}

\begin{proof}
  There are $\binom{m}{2}$ choices for the pair $(i, j)$ and $L^2$
  choices for $(\alpha, \beta)$.  By Lemma~\ref{lem:inject}, the
  $\binom{m}{2} L^2$ corresponding substrings are pairwise distinct
  binary strings, each beginning with~\texttt{1}, and hence represent
  pairwise distinct positive integers.  They are all substrings of $s_n$,
  so $f(s_n) \ge \binom{m}{2} L^2$ and thus $a(n) \ge \binom{m}{2} L^2$.
\end{proof}

Theorem~\ref{thm:main} now follows.

\begin{proof}[Proof of Theorem~\ref{thm:main}]
  The upper bound $\limsup_{n\to\infty} a(n)/n^2 \le \frac{1}{2}$ is
  Lemma~\ref{lem:upper}.  For the lower bound, use Lemma~\ref{lem:count}
  together with $L \ge (n - C)/m - 1 = (n - C - m)/m$:
  \[
    a(n)
    \;\ge\; \binom{m}{2} L^2
    \;\ge\; \frac{m(m-1)}{2} \cdot \left(\frac{n - C - m}{m}\right)^{\!2}
    \;=\; \frac{m-1}{2m}\,(n - C - m)^2.
  \]
  Dividing by $n^2$:
  \begin{equation}\label{eq:ratio}
    \frac{a(n)}{n^2}
    \;\ge\; \frac{1}{2}\!\left(1 - \frac{1}{m}\right)\!
            \left(1 - \frac{C + m}{n}\right)^{\!2}.
  \end{equation}
  Now $C + m = m(m-1)/2 + m = m(m+1)/2$.  Since $m = \floor{n^{1/3}}$,
  we have $m \to \infty$ and
  \[
    \frac{C + m}{n} = \frac{m(m+1)/2}{n} = O\!\left(\frac{m^2}{n}\right)
    = O\!\left(n^{-1/3}\right) \to 0.
  \]
  Therefore the right-hand side of~\eqref{eq:ratio} tends to
  $\frac{1}{2} \cdot 1 \cdot 1 = \frac{1}{2}$, giving
  $\liminf_{n\to\infty} a(n)/n^2 \ge \frac{1}{2}$.

  Combining with the upper bound completes the proof.
\end{proof}

% =======================================================================
\section{Quantitative corollary}\label{sec:quant}
% =======================================================================

\begin{corollary}
  For all $n \ge 8$,
  \[
    a(n) \;\ge\; \frac{n^2}{2} - O\!\left(n^{5/3}\right).
  \]
\end{corollary}

\begin{proof}
  Starting from $a(n) \ge \frac{m-1}{2m}(n - C - m)^2$, expand
  $(n - C - m)^2 = n^2 - 2n(C+m) + (C+m)^2$.  Since $C + m =
  m(m+1)/2 = O(m^2) = O(n^{2/3})$ and $m = O(n^{1/3})$:
  \begin{align*}
    (n - C - m)^2 &= n^2 - 2n \cdot O(n^{2/3}) + O(n^{4/3})
                   = n^2 - O(n^{5/3}),\\
    \frac{m-1}{2m} &= \frac{1}{2} - \frac{1}{2m} = \frac{1}{2} - O(n^{-1/3}).
  \end{align*}
  Thus
  \[
    a(n) \;\ge\; \left(\tfrac{1}{2} - O(n^{-1/3})\right)\bigl(n^2 - O(n^{5/3})\bigr)
         \;=\; \frac{n^2}{2} - O(n^{5/3}). \qedhere
  \]
\end{proof}

\begin{remark}
  The error term $O(n^{5/3})$ is an artefact of the simple construction.
  Computational experiments using suffix-array-guided local search
  yield strings achieving ratios within $0.004\%$ of $\frac{1}{2}$ at
  $n = 2\times 10^9$, far closer to the upper bound, but those
  refined constructions are not needed for the asymptotic theorem.
\end{remark}

% =======================================================================
\section{A worked example}\label{sec:example}
% =======================================================================

\begin{example}[$n = 27$]
  With $n = 27$ one computes $m = \floor{27^{1/3}} = 3$, $C = 3$,
  $L = \floor{24/3} = 8$, $r = 0$.  The construction gives
  \[
    s_{27} = \underbrace{11111111}_{8}\,0\,
             \underbrace{11111111}_{8}\,00\,
             \underbrace{11111111}_{8},
  \]
  a 27-bit string of the form $\texttt{1}^8\texttt{0}^1\texttt{1}^8
  \texttt{0}^2\texttt{1}^8$.

  The three pairs $(i,j)$ with $1 \le i < j \le 3$ are $(1,2)$, $(1,3)$,
  and $(2,3)$.  For each pair there are $L^2 = 64$ choices of $(\alpha,
  \beta)$, giving $3 \cdot 64 = 192$ distinct substrings:
  \begin{itemize}
    \item $(i,j) = (1,2)$: strings of the form
      $\texttt{1}^{\alpha}\texttt{0}\texttt{1}^{\beta}$
      (spanning the first separator).
    \item $(i,j) = (1,3)$: strings of the form
      $\texttt{1}^{\alpha}\texttt{0}\texttt{1}^{8}\texttt{00}\texttt{1}^{\beta}$
      (spanning both separators).
    \item $(i,j) = (2,3)$: strings of the form
      $\texttt{1}^{\alpha}\texttt{00}\texttt{1}^{\beta}$
      (spanning the second separator).
  \end{itemize}
  These three families are distinguished by their zero-block structure
  (one zero, then eight ones then two zeros, then two zeros,
  respectively), confirming injectivity directly.  Hence $a(27) \ge 192$.
  The exact value is $a(27) = 246$ (easily verified by exhaustive
  enumeration of all $2^{26}$ strings of length~$27$), so the bound is
  not tight but the asymptotic argument is correct.
\end{example}

% =======================================================================
\section*{Acknowledgements}
% =======================================================================

The author thanks the contributors to OEIS sequence A112509.

% =======================================================================
% Bibliography
% =======================================================================

\begin{thebibliography}{9}

\bibitem{oeis}
  OEIS Foundation Inc.,
  \emph{The On-Line Encyclopedia of Integer Sequences},
  Sequence A112509,
  \url{https://oeis.org/A112509}, 2026.

\bibitem{github}
  B.~Clare,
  \emph{A112509: computational exploration of OEIS sequence A112509},
  \url{https://github.com/axelsmagichammer/A112509}, 2026.

\end{thebibliography}

\end{document}
