\documentclass[11pt,a4paper]{article}

% -----------------------------------------------------------------------
% Packages
% -----------------------------------------------------------------------
\usepackage{amsmath,amsthm,amssymb,mathtools}
\usepackage{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{url}

\geometry{margin=1in}

% -----------------------------------------------------------------------
% Theorem environments
% -----------------------------------------------------------------------
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% -----------------------------------------------------------------------
% Macros
% -----------------------------------------------------------------------
\newcommand{\floor}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\val}{\nu}
\newcommand{\ones}{N_1}
\newcommand{\zeros}{N_0}

\title{Asymptotic and structural results for OEIS sequence A112509}
\author{Ben Clare}
\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
For a binary string $s$ of length $n$, let $f(s)$ be the number of
distinct non-negative integers represented by contiguous substrings of
$s$, read in binary with leading zeros suppressed.  Define
$a(n)=\max_{|s|=n}f(s)$ (OEIS A112509~\cite{oeis}).  The present paper
provides the first asymptotic results for this sequence.

We prove that for all $n\ge 64$,
\[
  a(n) \;\ge\; \tfrac{1}{2}n^2 - 15\,n^{3/2},
\]
and hence $\lim_{n\to\infty}a(n)/n^2=\tfrac{1}{2}$, matching the
elementary upper bound $a(n)\le n(n+1)/2$.  The proof is completely
explicit.  Beyond the limit, we prove:\ (i)~near-optimal strings can
be chosen with ones-density $1-O(n^{-1/2})$;\ (ii)~every optimal
string has ones-density $1-O(n^{-1/4})$;\ (iii)~the number of zeros
$z$ in any string $s$ reduces the maximum possible $f(s)$ by at least
$z(z+1)/2$;\ (iv)~fragmenting the zeros into $m$ runs imposes an
additional penalty of at least $\binom{m}{2}$.  Computational data for
$n\le150$ and certified lower bounds through $n=2\times10^9$ are
discussed~\cite{github}.
\end{abstract}

\section{Introduction}

\subsection{The sequence}

For $s\in\{0,1\}^n$, let $f(s)$ denote the number of distinct
non-negative integers whose binary representations appear as contiguous
substrings of $s$ (leading zeros suppressed: $\mathtt{01}=1$,
$\mathtt{00}=0$).  Set $a(n)=\max_{|s|=n}f(s)$.  This is OEIS
A112509~\cite{oeis}, beginning
\[
  1,3,5,7,10,13,17,22,27,33,40,47,55,64,73,83,94,106,118,131,\ldots
\]

\subsection{Numerical evidence}

Exact values for $n\le150$ were computed in~\cite{github} by
exhaustive structured search, and certified lower bounds obtained for
much larger $n$ via suffix-array-guided local search.
Table~\ref{tab:evidence} records selected values.

\begin{table}[ht]
\centering
\begin{tabular}{rrl}
\toprule
$n$ & $a(n)$ or lower bound & $a(n)/n^2\ge$ \\
\midrule
$50$            & $899$                                    & $0.3596$ \\
$100$           & $3{,}875$                                & $0.3875$ \\
$150$           & $9{,}070$                                & $0.4031$ \\
$10^6$          & $498{,}396{,}662{,}200$                  & $0.4984$ \\
$10^9$          & $499{,}943{,}911{,}277{,}037{,}650$      & $0.49994$ \\
$2\times10^9$   & $1{,}999{,}841{,}302{,}297{,}432{,}600$ & $0.49996$ \\
\bottomrule
\end{tabular}
\caption{Selected values and lower bounds for $a(n)$.}
\label{tab:evidence}
\end{table}

\subsection{Main results}

\begin{theorem}[Limit]\label{thm:limit}
$\displaystyle\lim_{n\to\infty}\dfrac{a(n)}{n^2}=\dfrac{1}{2}$.
\end{theorem}

\begin{theorem}[Quantitative lower bound]\label{thm:quant}
For every $n\ge64$,
\[
  a(n)\;\ge\;\frac{1}{2}n^2-15\,n^{3/2}.
\]
\end{theorem}

\begin{proposition}[Density of ones]\label{prop:density}
\leavevmode
\begin{enumerate}[label=\normalfont(\roman*)]
\item There exists an explicit family of strings $s_n$ with
  $f(s_n)\ge\tfrac{1}{2}n^2-15n^{3/2}$ and $\zeros(s_n)=O(\sqrt{n})$,
  so the ones-density satisfies $\ones(s_n)/n=1-O(n^{-1/2})$.
\item Every optimal string $s_n^*$ satisfies $\zeros(s_n^*)=O(n^{3/4})$,
  equivalently $\ones(s_n^*)/n=1-O(n^{-1/4})$.
\end{enumerate}
\end{proposition}

Section~\ref{sec:upper} gives the upper bound.
Sections~\ref{sec:construction}--\ref{sec:lower} prove
Theorems~\ref{thm:limit}--\ref{thm:quant}.
Section~\ref{sec:density} proves Proposition~\ref{prop:density}.
Sections~\ref{sec:zeros}--\ref{sec:runs} develop the structural bounds.

\section{Definitions}

If $w=w_1\cdots w_k\in\{0,1\}^k$, its binary value is
$\val(w)=\sum_{i=1}^k w_i\,2^{k-i}$.
The value set of $s\in\{0,1\}^n$ is
$V(s)=\{\val(s_i\cdots s_j):1\le i\le j\le n\}$,
and $f(s)=|V(s)|$.
Write $\ones(s)$ and $\zeros(s)$ for the number of ones and zeros in $s$.

\section{Upper bound}\label{sec:upper}

\begin{lemma}[Upper bound]\label{lem:upper}
For every $n\ge1$, $a(n)\le n(n+1)/2$, and hence
$\limsup_{n\to\infty}a(n)/n^2\le1/2$.
\end{lemma}

\begin{proof}
Every distinct positive integer in $V(s)$ has a unique standard binary
representation beginning with~$\mathtt{1}$, so it is represented by at
least one substring of $s$ beginning with~$\mathtt{1}$.  Different
positive integers require different such substrings.

If $s$ contains at least one zero then $0\in V(s)$, represented only by
substrings beginning with~$\mathtt{0}$.  At least one of the $n(n+1)/2$
substrings begins with~$\mathtt{0}$, so the number of leading-$\mathtt{1}$
substrings is at most $n(n+1)/2-1$, giving $f(s)\le n(n+1)/2$.  If $s$
is all ones then $f(s)$ counts distinct positive values, at most $n(n+1)/2$.
\end{proof}

\begin{remark}
The bound is tight at $n=2$: $\mathtt{10}$ gives $V(\mathtt{10})=\{0,1,2\}$,
so $a(2)=3=2\cdot3/2$.
\end{remark}

\section{The construction}\label{sec:construction}

\subsection{Strictly decreasing partitions}

\begin{lemma}[Decreasing partition]\label{lem:partition}
Let $k\ge2$ and $M\ge k(k+1)/2$.  Then there exist strictly decreasing
positive integers $b_1>b_2>\cdots>b_k\ge1$ with $\sum_{i=1}^k b_i=M$.
Moreover, they may be chosen with $|b_i-M/k|\le k$ for all $i$.
\end{lemma}

\begin{proof}
\textit{Strategy.}  Consider the arithmetic sequence $c_i=q+k-i$
($i=1,\ldots,k$), which decreases by~$1$ at each step.  Its sum is
\begin{equation}\label{eq:basesum}
  \sum_{i=1}^{k}c_i
  = kq + \sum_{i=1}^{k}(k-i)
  = kq + \frac{k(k-1)}{2}.
\end{equation}
Choose $q$ so this base sum is as large as possible without exceeding
$M$, and let $r=M-\sum c_i$ be the shortfall.  We then add $1$ to the
first $r$ terms to account for the shortfall.

\textit{Definitions.}  Set
\[
  q = \Bigl\lfloor \frac{M}{k} - \frac{k-1}{2} \Bigr\rfloor,
  \qquad
  r = M - kq - \frac{k(k-1)}{2},
\]
and define
\[
  b_i=\begin{cases}q+k-i+1,&1\le i\le r,\\q+k-i,&r<i\le k.\end{cases}
\]

\textit{Step~1: $0\le r<k$.}
From the definition of $q$ as a floor, $q\le M/k-(k-1)/2<q+1$.
Multiplying through by $k$ gives
\[
  kq \;\le\; M - \tfrac{k(k-1)}{2} \;<\; k(q+1),
\]
so $0\le M-kq-k(k-1)/2<k$, i.e.\ $0\le r<k$.

\textit{Step~2: Strict decrease.}
We check all three cases for consecutive pairs $b_i>b_{i+1}$.
\begin{itemize}[nosep]
  \item \textit{Both in the first case} ($i<r$, so $i+1\le r$ as well):
    $b_i-b_{i+1}=(q+k-i+1)-(q+k-(i+1)+1)=1>0$.
  \item \textit{Transition} ($i=r\ge1$, so $i+1>r$):
    $b_r=q+k-r+1$ and $b_{r+1}=q+k-(r+1)$,
    giving $b_r-b_{r+1}=q+k-r+1-(q+k-r-1)=2>0$.
  \item \textit{Both in the second case} ($i>r$):
    $b_i-b_{i+1}=(q+k-i)-(q+k-i-1)=1>0$.
\end{itemize}
Hence $b_1>b_2>\cdots>b_k$.

\textit{Step~3: The sum equals $M$.}
Split the sum at the boundary between the two cases:
\begin{align*}
  \sum_{i=1}^{k}b_i
  &= \sum_{i=1}^{r}(q+k-i+1) + \sum_{i=r+1}^{k}(q+k-i) \\
  &= \sum_{i=1}^{k}(q+k-i) + \sum_{i=1}^{r}1 \\
  &= kq + \frac{k(k-1)}{2} + r \;=\; M,
\end{align*}
where the last equality uses the definition of~$r$ and~\eqref{eq:basesum}.

\textit{Step~4: All terms are positive.}
From $M\ge k(k+1)/2$ we get $M/k\ge(k+1)/2$, so
\[
  \frac{M}{k}-\frac{k-1}{2}\;\ge\;\frac{k+1}{2}-\frac{k-1}{2}=1,
\]
hence $q=\lfloor M/k-(k-1)/2\rfloor\ge1$.
Since $r<k$, the index $i=k$ falls in the second case, giving
$b_k=q+k-k=q\ge1$.

\textit{Step~5: Proximity bound.}
Every first-case term satisfies $q < q+k-r+1\le b_i\le q+k$, and
every second-case term satisfies $q\le b_i=q+k-i\le q+k-r-1<q+k$.
So all $b_i\in[q,\,q+k]$.
From the floor definition, $M/k\in[q+(k-1)/2,\;q+(k+1)/2)$,
and the width of $[q,q+k]$ around any point in this interval is at most
$(k+1)/2<k$ in each direction.
Hence $|b_i-M/k|<k$ for all~$i$.
\end{proof}

\subsection{Parameters and string}

Fix $n\ge64$ and set
\begin{equation}\label{eq:params}
  k=\floor{\tfrac{\sqrt{n}}{4}},\qquad M=n-(k-1).
\end{equation}
Since $n\ge64$ we have $k=\lfloor\sqrt{n}/4\rfloor\ge\lfloor8/4\rfloor=2$.

\textit{Why $M\ge n/2$.}
Since $k\le\sqrt{n}/4$, we have $\sqrt{n}/4\le n/2$ (equivalently $\sqrt{n}\le2n$, true for all $n\ge1$),
so $k-1<k\le\sqrt{n}/4\le n/2$.
Hence $M=n-(k-1)>n-n/2=n/2$.

\textit{Why $n/2\ge k(k+1)/2$.}
This is equivalent to $n\ge k(k+1)$.  Since $k\le\sqrt{n}/4$ we have
\[
  k(k+1)\le k\cdot2k=2k^2\le2\cdot\frac{n}{16}=\frac{n}{8}\le n,
\]
where the second inequality uses $k\le\sqrt{n}/4$ so $k^2\le n/16$.
Hence $M\ge n/2\ge k(k+1)/2$, and Lemma~\ref{lem:partition} applies.

Choose $b_1>b_2>\cdots>b_k\ge1$ with $\sum b_i=M$ and
$|b_i-M/k|\le k$, as in Lemma~\ref{lem:partition}.  Define
\begin{equation}\label{eq:string}
  s\;=\;\underbrace{1^{b_1}}_{\text{block }1}\,0\,
         \underbrace{1^{b_2}}_{\text{block }2}\,0\,
         \cdots\,0\,
         \underbrace{1^{b_k}}_{\text{block }k},
\end{equation}
consisting of $k$ one-blocks separated by $k-1$ single zeros.  Its
length is $\sum b_i+(k-1)=M+(k-1)=n$.

\section{An injective family of substrings}\label{sec:family}

For integers $i,j,\alpha,\beta$ satisfying
\begin{equation}\label{eq:range}
  1\le i,\quad j\ge i+2,\quad j\le k,\quad
  1\le\alpha\le b_i,\quad 1\le\beta\le b_j,
\end{equation}
define $u_{i,j,\alpha,\beta}$ to be the substring of $s$ beginning
$\alpha$ positions before the end of block~$i$ and ending after the
first $\beta$ ones of block~$j$.  In binary,
\begin{equation}\label{eq:u}
  u_{i,j,\alpha,\beta}
  =1^{\alpha}\,0\,1^{b_{i+1}}\,0\,1^{b_{i+2}}\,0\,
  \cdots\,1^{b_{j-1}}\,0\,1^{\beta}.
\end{equation}
The condition $j\ge i+2$ ensures at least one full interior block
($b_{i+1},\ldots,b_{j-1}$).  Every $u_{i,j,\alpha,\beta}$ starts and
ends with~$\mathtt{1}$, so it is already a standard binary expansion of
a positive integer.

\begin{lemma}[Injectivity]\label{lem:inject}
The map $(i,j,\alpha,\beta)\mapsto u_{i,j,\alpha,\beta}$ is injective
on the range~\eqref{eq:range}.
\end{lemma}

\begin{proof}
Suppose $u_{i,j,\alpha,\beta}=u_{i',j',\alpha',\beta'}$ as binary words.
The run-length encoding (RLE) of~\eqref{eq:u}, listing alternating
run-lengths of ones then zeros, is
\[
  (\alpha,\,1,\,b_{i+1},\,1,\,b_{i+2},\,1,\,\ldots,\,b_{j-1},\,1,\,\beta).
\]
Matching entries gives $\alpha=\alpha'$, $\beta=\beta'$, and
$(b_{i+1},\ldots,b_{j-1})=(b_{i'+1},\ldots,b_{j'-1})$.  Since
$b_1>\cdots>b_k$ is strictly decreasing, every consecutive subsequence
occurs in exactly one position, so $i=i'$ and $j=j'$.
\end{proof}

\section{Quantitative lower bound and the limit}\label{sec:lower}

\begin{proof}[Proof of Theorem~\ref{thm:quant}]
By Lemma~\ref{lem:inject}, the family $\{u_{i,j,\alpha,\beta}\}$
consists of pairwise distinct positive integers, so
\begin{equation}\label{eq:f-lower}
  a(n)\ge f(s)\ge\sum_{\substack{1\le i<j\le k\\j\ge i+2}}b_i b_j.
\end{equation}
Expanding:
\begin{equation}\label{eq:expand}
  \sum_{j\ge i+2}b_i b_j
  =\frac{(\sum b_i)^2-\sum b_i^2}{2}-\sum_{i=1}^{k-1}b_i b_{i+1}.
\end{equation}

\textbf{Step~1.}  Since $\sum b_i=M=n-k+1$ and $k\le\sqrt{n}/4$,
\[
  \frac{M^2}{2}\ge\frac{(n-\sqrt{n}/4)^2}{2}
  \ge\frac{n^2}{2}-\frac{n^{3/2}}{4}.
\]

\textbf{Step~2.}  Write $b_i=M/k+d_i$ with $|d_i|\le k$.  Then
$\sum b_i^2\le M^2/k+k^3$.
Since $k\ge\sqrt{n}/8$ (as $\lfloor x\rfloor\ge x/2$ for $x\ge2$),
$M^2/k\le n^2/(\sqrt{n}/8)=8n^{3/2}$,
and $k^3\le(\sqrt{n}/4)^3=n^{3/2}/64$, so
\begin{equation}\label{eq:sumsq}
  \sum_{i=1}^k b_i^2<9n^{3/2}.
\end{equation}

\textbf{Step~3.}  By AM--GM,
$b_i b_{i+1}\le(b_i^2+b_{i+1}^2)/2$, so
\begin{equation}\label{eq:adj}
  \sum_{i=1}^{k-1}b_i b_{i+1}\le\sum_{i=1}^k b_i^2<9n^{3/2}.
\end{equation}

\textbf{Conclusion.}  Substituting into~\eqref{eq:expand}:
\[
  a(n)\ge\frac{n^2}{2}-\frac{n^{3/2}}{4}
  -\frac{9n^{3/2}}{2}-9n^{3/2}
  =\frac{n^2}{2}-\frac{55}{4}n^{3/2}>\frac{n^2}{2}-15n^{3/2}.\qedhere
\]
\end{proof}

\begin{proof}[Proof of Theorem~\ref{thm:limit}]
Lemma~\ref{lem:upper} gives $\limsup a(n)/n^2\le1/2$.
Theorem~\ref{thm:quant} gives $a(n)/n^2\ge1/2-15/\sqrt{n}\to1/2$.
\end{proof}

\begin{corollary}\label{cor:rate}
For all $n\ge64$, $\;n^2/2-15n^{3/2}\le a(n)\le n(n+1)/2$, so
$a(n)/n^2=1/2+O(n^{-1/2})$.
\end{corollary}

\begin{remark}
The lower bound $n^2/2-15n^{3/2}$ is negative for $n<900$ (since
$n^2/2>15n^{3/2}$ iff $\sqrt{n}>30$).  For such $n$ the inequality in
Theorem~\ref{thm:quant} is trivially satisfied by the non-negativity of
$a(n)$, and carries no information.  The bound is meaningful only for
$n\ge900$; for small~$n$ the exact values from computational search
(Table~\ref{tab:values}) give much sharper information.
\end{remark}

\section{Density of ones}\label{sec:density}

\subsection{Near-optimal strings}

\begin{proof}[Proof of Proposition~\ref{prop:density}(i)]
The string $s=s_n$ from~\eqref{eq:string} has $f(s_n)\ge n^2/2-15n^{3/2}$
by Theorem~\ref{thm:quant}, and contains exactly $k-1\le\sqrt{n}/4$
zeros, so $\zeros(s_n)=O(\sqrt{n})$ and $\ones(s_n)/n=1-O(n^{-1/2})$.
\end{proof}

\subsection{Zero-position penalty}

\begin{lemma}[Zero-position bound]\label{lem:zero-pos}
Let $s\in\{0,1\}^n$ have zeros at positions $p_1<\cdots<p_z$.  Then
\[
  f(s)\le\frac{n(n+1)}{2}-\sum_{r=1}^z(n-p_r+1).
\]
In particular, $\;f(s)\le n(n+1)/2-z(z+1)/2$.
\end{lemma}

\begin{proof}
For each $x\in V(s)$ choose a minimum-length representative $w_x$.
The map $x\mapsto w_x$ is injective.  If $x>0$ then $w_x$ starts with
$\mathtt{1}$ (otherwise removing leading zeros gives a shorter
representative).  So distinct positive values correspond to distinct
leading-$\mathtt{1}$ substrings.

Substrings starting at zero-position $p_r$ number $n-p_r+1$ and none
starts with~$\mathtt{1}$, giving the first inequality.  The second
follows by placing zeros as far right as possible ($p_r=n-z+r$), which
minimises $\sum(n-p_r+1)=z(z+1)/2$.
\end{proof}

\subsection{Optimal strings have few zeros}

\begin{proof}[Proof of Proposition~\ref{prop:density}(ii)]
Let $s_n^*$ be optimal with $z_n=\zeros(s_n^*)$.  By
Theorem~\ref{thm:quant} and Lemma~\ref{lem:zero-pos},
\[
  \frac{n^2}{2}-15n^{3/2}
  \le a(n)=f(s_n^*)
  \le\frac{n(n+1)}{2}-\frac{z_n(z_n+1)}{2}.
\]
Hence $z_n(z_n+1)/2\le15n^{3/2}+n/2+1$, giving $z_n=O(n^{3/4})$ and
$\ones(s_n^*)/n=1-O(n^{-1/4})$.
\end{proof}

\section{Upper bounds via the zero structure}\label{sec:zeros}

Lemma~\ref{lem:zero-pos} treats all zeros equally.  We now refine the
bound using the \emph{run structure} of the zeros.

Let the maximal zero-runs of $s$ have lengths $c_1,\ldots,c_m$, with
the $i$-th run beginning at position $a_i$ and $r_i=n-(a_i+c_i)+1$
positions to its right.

\begin{lemma}[Run-sensitive bound]\label{lem:runs-raw}
$f(s)\le n(n+1)/2-\sum_{i=1}^m\bigl(c_i r_i+\binom{c_i+1}{2}\bigr)$.
\end{lemma}

\begin{proof}
Apply Lemma~\ref{lem:zero-pos} and group by zero-runs.
The $i$-th run occupies positions $a_i,\ldots,a_i+c_i-1$, contributing
$\sum_{t=0}^{c_i-1}(n-a_i-t+1)=c_i(n-a_i+1)-c_i(c_i-1)/2
=c_i r_i+\binom{c_i+1}{2}$ to the sum.
\end{proof}

\section{Fragmentation penalty}\label{sec:runs}

\begin{corollary}[Fragmentation]\label{cor:fragment}
If $s$ has $z$ zeros in $m$ maximal zero-runs, then
\[
  f(s)\;\le\;\frac{n(n+1)}{2}-\frac{z(z+1)}{2}-\binom{m}{2}.
\]
\end{corollary}

\begin{proof}
It suffices to show
$\sum_i\bigl(c_i r_i+\binom{c_i+1}{2}\bigr)\ge z(z+1)/2+\binom{m}{2}$.
Since consecutive zero-runs are separated by at least one $\mathtt{1}$,
for $i<m$ we have $r_i\ge\sum_{j>i}c_j+(m-i)$, so
$\sum_i c_i r_i\ge\sum_{i<j}c_i c_j+\sum_{i=1}^{m-1}c_i(m-i)$.
The identity $\binom{z+1}{2}=\sum_{i<j}c_i c_j+\sum_i(c_i^2+c_i)/2$ gives
$\sum_{i<j}c_i c_j+\sum_i\binom{c_i+1}{2}=z(z+1)/2$.
Finally $\sum_{i=1}^{m-1}c_i(m-i)\ge\sum_{i=1}^{m-1}(m-i)=\binom{m}{2}$
(each $c_i\ge1$).
\end{proof}

\begin{remark}
Corollary~\ref{cor:fragment} shows that fragmenting the zeros into more
runs always reduces the upper bound on $f(s)$, regardless of positions.
Each additional run incurs an extra penalty of at least $m-1$ (the new
run number), on top of the baseline $z(z+1)/2$.
\end{remark}

\section{Computational evidence}\label{sec:comp}

\paragraph{Exact values ($n\le150$).}
Exact values of $a(n)$ were computed for all $n\le150$ in~\cite{github}
using three independent methods: unconstrained Metropolis--Hastings
search (at most 10~restarts, reaching the optimum for every $n\le150$),
constrained Metropolis--Hastings with a fixed run-length template, and
exact exhaustive template search.  On every overlap the methods agreed.

\paragraph{Large-scale certified lower bounds.}
A suffix-array-guided local search (\emph{surgical nudge}) was used for
$n$ up to $2\times10^9$~\cite{github}.  In each iteration the algorithm
finds the highest-LCP collision in the suffix array, searches a
neighbourhood for a density-preserving swap that resolves it, and
accepts only swaps that strictly increase the exact score.  Since the
score is monotone non-decreasing, every recorded value is a rigorous
lower bound.  At $n=2\times10^9$ a lower bound of $a(n)/n^2\ge0.49996$
was obtained.

\paragraph{Comparison with theory.}
The ratio $a(n)/n^2$ rises from $0.3596$ at $n=50$ to $0.4031$ at
$n=150$, and exceeds $0.49996$ at $n=2\times10^9$.  
Theorem~\ref{thm:quant} gives $a(n)/n^2\ge1/2-15/\sqrt{n}$, which
reaches $0.4875$ only at $n=14{,}400$.  The computational lower bounds
substantially exceed the theoretical bound at moderate $n$, confirming
that the constant $15$ in $15n^{3/2}$ is far from tight.

\paragraph{Entropy of optimal solutions.}
Let $S_n$ denote the set of all strings achieving $a(n)$ and
$H(n)=\log_2|S_n|$.  For $n\le150$, $H(n)$ is almost always zero or
very small: the typical optimum is unique or nearly so, with
$H(n)\le10.4$ throughout.  Optimal strings share long common prefixes;
the zero-separator structure is completely rigid in 28~cases and varies
by at most one separator in a further 119~cases out of $n\le150$.

\section{Heuristic remarks}\label{sec:heuristic}

To maximise $f(s)$, one seeks to maximise the number of
leading-$\mathtt{1}$ substrings that remain pairwise distinct after
leading zeros are stripped.  Long repeated patterns create large
common-prefix collisions, inflating LCP values and causing many pairs
of substrings to agree.  The explicit construction in
Section~\ref{sec:construction} avoids this by using \emph{strictly
decreasing} one-block lengths: any two substrings $u_{i,j,\alpha,\beta}$
and $u_{i',j',\alpha',\beta'}$ spanning different block-pairs $(i,j)
\ne(i',j')$ have different interior block-length sequences, making them
provably distinct (Lemma~\ref{lem:inject}).

Empirically, all optimal strings for $n\le150$ follow a similar
principle: they begin with a long run of ones, followed by
decreasing-length one-blocks with a fixed separator prefix
$1,2,1,1,\ldots$, and have ones-density growing toward $1$.  This
structure is consistent with the proven density bounds, though the
actual empirical zero count at small $n$ grows far more slowly than the
theoretical $O(n^{3/4})$ bound would suggest.

\section{A worked example}\label{sec:example}

\begin{example}[$n=169$, $k=3$]
With $n=169$: $k=\lfloor13/4\rfloor=3$, $M=169-2=167$.  Choose
\[
  b_1=57,\quad b_2=56,\quad b_3=54;\qquad
  57>56>54\ge1,\quad57+56+54=167\,.
\]
The construction gives the 169-bit string
\[
  s=\underbrace{1^{57}}_{\text{block }1}\;0\;
    \underbrace{1^{56}}_{\text{block }2}\;0\;
    \underbrace{1^{54}}_{\text{block }3},
\]
of length $57+1+56+1+54=169$.

The only valid pair with $j\ge i+2$ and $j\le k=3$ is $(i,j)=(1,3)$,
with interior block $b_2=56$.  For each $(\alpha,\beta)$ with
$1\le\alpha\le57$ and $1\le\beta\le54$, the substring
$u_{1,3,\alpha,\beta}=1^{\alpha}\,0\,1^{56}\,0\,1^{\beta}$.
By Lemma~\ref{lem:inject} these $57\times54=3{,}078$ substrings are
pairwise distinct positive integers, so $a(169)\ge3{,}078$.

Note: the explicit bound $n^2/2-15n^{3/2}<0$ at $n=169$, reflecting
that the constant $15$ is conservative and the asymptotic result
requires large $n$ to dominate.
\end{example}

\section*{Acknowledgements}

The author thanks the contributors to OEIS sequence A112509.

\begin{thebibliography}{9}

\bibitem{oeis}
  OEIS Foundation Inc.,
  \emph{The On-Line Encyclopedia of Integer Sequences},
  Sequence A112509,
  \url{https://oeis.org/A112509}, 2026.

\bibitem{github}
  B.~Clare,
  \emph{A112509: computational exploration of OEIS sequence A112509},
  \url{https://github.com/axelsmagichammer/A112509}, 2026.

\end{thebibliography}

\end{document}
